Complete Algebraic File Synchronization

### 2015-12-02 2015/paper.tex
Proof for the completeness of reconciliation. Only one more proof to go!

\newcommand{\typeeq}{\stackrel{t}{=}}
\newcommand{\ntypeeq}{\centernot{\stackrel{t}{=}}}

@@

@@

\newtheorem{myax}{Axiom}
\crefname{myax}{axiom}{axioms}

@@

\title{The Correctness of Algebraic File Synchronization}

\author{Elod Pal Csirmaz}

@@

@@

In this paper, we write $\valf$ for an arbitrary element of $\setf$, and $\vald$ for an arbitrary element
of $\setd$. $TODO$ is a value from $\setv$ and $n$ and $m$ are nodes in $\setn$.

@@

\section{Investigating command pairs}

@@

Let $A$ be a sequence that satisfies the conditions;
we therefore know that it contains a command $\cxynv$ on $n$
and another command $\czwnnv$ on $\nn$.
We use an indirect proof and assume that there are no commands on $\parent(n)$ in $A$.
Next, we create a new sequence $A'\equiv A$ in which $\cxy$ and $\czw$ are next to each other.
If they are already next to each other in $A$, there is nothing to do.
Otherwise, consider the command next to $\cxynv$ in the direction where $\czwnnv$ is.
Let this command be $\cqrmv$.
If $\czw$ is to the right, then $A$ looks like the following:
\[ A = \cdots\cc\cxynv\cc\cqrmv\cc\cdots\cc\czwnnv\cc\cdots \]
If $n\unrel m$, swap $\cxynv$ and $\cqrmv$. Based on \cref{ax_separate_commute} we know that the new
sequence is equivalent to $A$.
Otherwise, we know $m\neq\parent(n)$ as there are no commands on $\parent(n)$, and so
from \cref{ax_distantrel_breaks} we get $A\equiv\cbrk$ which contradicts our assumptions.
(Note that $A$ cannot contains assertion commands.)
By repeating this step we can therefore convert $A$ into $A'$ where $\cxynv$ and $\czwnnv$ are neighboring commands.
However, then \cref{ax_distantrel_breaks} applies to the sequence and therefore $A\equiv A'\equiv\cbrk$ which
is again a contradiction.
\end{proof}

@@

To order the commands in the component,
select the command with the topmost node, so that there would be no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $X\empt$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any sequence to break all filesystems.

@@

Based on \cref{ax_same_emptyseq,ax_same_singlec} there is
an $S''\eqnrw S'$ in which multiple commands on the same nodes
are simplified into a single command or an empty sequence,
and therefore $S''$ is simple and $S\equiv S'\eqext S''$.
It is easy to see that $S''$ must contain the same commands
as the output of the update detector, $U$, as both are simple
but reflect all changes between $\FS$ and $\FS^*$.
If so, then based on \cref{simple_reorder_equiv}
we have $S''\in\orderset{U}$, from which
$S\equiv S'\eqext S''\equiv U$, and as $S\FS = \FS^* \neq \fsbroken$,
therefore $U\FS\neq\fsbroken$.
\end{proof}

@@

\begin{mydef}{Independent commands, sequences and sets.}
For two commands $\cxynv$ and $\czwmv$ we write $\cxynv\indep\czwmv$ if $\cxynv\cc\czwmv \equiv \czwnv\cc\cxynv \nequiv \cbrk$.
For two sequences or sets of commands $A$ and $B$ we write $A\indep B$ if
$\forall a\in A \mbox{~and~} \forall b\in B: a\indep b$.
\end{mydef}

\begin{mydef}{Reconciliation.}\label{def:reconciliation}
\begin{align*}
\reca = \{c| & \\
& c\in A\setminus B \\
& \wedge \\
& \forall c'\in B\setminus A: c\indep c'\\
\} &
\end{align*}

In other words, the sequence $\reca$ we apply to $\FS_B$ consists of all commands from $A\setminus B$
that are independent of every command in $B\setminus A$.
The order of commands in the sequence can be determined using the algorithm described in
\cref{ordering} as any subset of $A$ is also simple.
In a similar way, $\recb$ can be obtained by reversing $A$ and $B$.
\end{mydef}

@@

\begin{mylem}\label{unrelated_is_independent}
Two different commands that are not assertion commands are independent iff the nodes they change are unrelated:
\[ \cxynv\indep\czwmv \Leftrightarrow n\unrel m \]
if $\cxynv\neq\czwmv$ and none of $\cxy$ and $\cxw$ is $\cbb$ or $\cdd$.
\end{mylem}
\begin{proof}
We proceed in two steps; first, we prove that
\[ \cxynv\nindep\czwmv \wedge n\unrel m \] would lead to contradiction.
By definition, if the two commands are not independent, then at least one of the following holds:
\begin{itemize}
\item $\czwmv\cc \cqrov \equiv\cbrk$
\item $\cqrov\cc \czwmv \equiv\cbrk$
\item $\czwmv\cc \cqrov \nequiv \cqrov\cc \czwmv$
\end{itemize}
However, from \cref{ax_separate_commute,ax_separate_nobreaks} we also know that
if $n\unrel m$ then none of the above can be true.

Next, we prove that
\[ \cxynv\indep\czwmv \wedge n\nunrel m \] would lead to contradiction.
As $n\nunrel m$
we know that $n=m$, $n\descendant m$ or $m\descendant n$.

@@

\begin{myth}
If $A, B$ are simple, then $\worksc{B\cc \reca}{A,B}$,
where, to restate \cref{def:reconciliation},
% { % scope
% \renewcommand{\themydef}{\ref{def:reconciliation}}
% \begin{mydef}
\begin{align*}
\reca = \{c| & \\
& c\in A\setminus B \\
& \wedge \\
& \forall c'\in B\setminus A: c\indep c'\\
\}. &
\end{align*}
% \end{mydef}
% } % scope
\end{myth}

@@

In these cases, either $m=\parent(n)$ or $n=\parent(m)$.
If $B$ has a command on $m$, then
from \cref{lemma:neighbor}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxy$ (also in $B$),
and from \cref{simple_reorder_equiv} we know that $B'\equiv B$.
When describing our ordering algorithm in \cref{ordering} we saw
that the command on the parent path deremines whether a pair of commands
is a construction or destruction pair,
and that this, in turn, determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $m$ must be on
the same side of the command on $n$ in both sequences.

If $B$ has no command on $m$, then let $B'$ be $B$, 
but with an extra command added to it just before $\cxy$
according to \cref{ax_child_assert,ax_parent_assert}, 
from which we know that $B'\equiv B$.
We also know that $B'$ is still minimal (but no longer simple).

@@

Going back to the construction and destruction pairs we see that the output type of the first command
is always the same, and it is either $D$ or $\empt$ depending on the relationship between $n$ and $m$.
Therefore $R=W$ and $\cqrmv=\czwmv$
as $|\setd|=|\setb|=1$. 
If $B$ originally had a command on $m$,
this is a contradiction as $\czw$ was not marked, but we see it must also be in $B'$ and therefore in $B$.
If $B$ had no command on $m$,
this is a contradiction because the command on $m$ in $B'$ is an assertion command, so $Z=Q=R=W$, 
but $A$ contained no assertion commands.

@@

\newcommand{\acb}{(A\cap B)}
\newcommand{\acbi}{\acb^{-1}}
\newcommand{\amb}{(A\setminus B)}
\newcommand{\bma}{(B\setminus A)}
We would like to prove that $\worksc{B\cc \reca}{A,B}$.
From the above we know that this is equivalent to
\[ \worksc{\acb\cc \bma\cc \reca}{A,B}. \]
As we already know that $\reca\indep\bma$
and that $\worksc{\acb\cc \bma}{A,B}$
(as $\worksc{B}{A,B}$),
based on \cref{indep_prefix_combine}
we only need to prove that
\[ \worksc{\acb\cc \reca}{A,B}. \]
Using \cref{r_invmove} we can rephrase this as
\[ \worksc{\reca}{\acbi\cc A,\acbi\cc B}. \]

@@

% Reconciliation is maximal
% -------------------------
% This is where we're using |D|=1
% C.f. strict / liberal etc. reconciliation in references

% TODO
TODO: $\reca$ is maximal

@@

% TODO
TODO: Are these results quite obvious as this is a well-known kind of algebra or system?

>>>>

\newcommand{\typeeq}{\stackrel{t}{=}} % Type equality
\newcommand{\ntypeeq}{\centernot{\stackrel{t}{=}}}

@@

\newcommand{\acb}{(A\cap B)}
\newcommand{\acbi}{\acb^{-1}}
\newcommand{\amb}{(A\setminus B)}
\newcommand{\bma}{(B\setminus A)}

@@

\newtheorem{myax}{Rule}
\crefname{myax}{rule}{rules}

@@

\title{Complete Algebraic File Synchronization}
\author{Elod Pal Csirmaz}

@@

% TODO Thank-you Laszlo Csirmaz, Bill Zissimopoulos

@@

In this paper, 
we write $n$ and $m$ for nodes in $\setn$,
and use lowercase Greek letters to denote commands. % TODO but very rarely
We use $\valf$ for an arbitrary element of $\setf$, and $\vald$ for an arbitrary element
of $\setd$. $TODO$ is a value from $\setv$ .

@@

\section{Investigating command pairs}\label{section_axioms}

@@

Let $A$ be a sequence that satisfies the conditions;
we therefore know that it contains a command $\cxynv$ on $n$
and another command $\czwnnv$ on $\nn$.
We use an indirect proof and assume that there are no commands on $\parent(n)$ in $A$.
Next, we create a new sequence $A'\equiv A$ in which $\cxynv$ and $\czwnnv$ are next to each other.
If they are already next to each other in $A$, there is nothing to do.
Otherwise, consider the command next to $\cxynv$ in the direction where $\czwnnv$ is.
Let this command be $\cqrmv$.
If $\czwnnv$ is to the right, then $A$ looks like the following:
\[ A = \cdots\cc\cxynv\cc\cqrmv\cc\cdots\cc\czwnnv\cc\cdots \]
If $n\unrel m$, swap $\cxynv$ and $\cqrmv$. Based on \cref{ax_separate_commute} we know that the new
sequence is equivalent to $A$.
Otherwise, we know $m\neq\parent(n)$ as there are no commands on $\parent(n)$, and so
from \cref{ax_distantrel_breaks} we get $A\equiv\cbrk$ which contradicts our assumptions.
(Note that $A$ cannot contains assertion commands.)
By repeating this step we can therefore convert $A$ into $A'$ where $\cxynv$ and $\czwnnv$ are neighboring commands.
However, then \cref{ax_distantrel_breaks} applies to the sequence and therefore $A\equiv A'\equiv\cbrk$ which
is again a contradiction.
\end{proof}

@@

To order the commands in the component,
select the command with the topmost node, so that there would be no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $\caa{X}{\ccharb}$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any sequence to break all filesystems.

@@

Based on \cref{ax_same_emptyseq,ax_same_singlec} there is
an $S''\eqnrw S'$ in which multiple commands on the same nodes
are simplified into a single command or an empty sequence,
and therefore $S''$ is simple and $S\equiv S'\eqext S''$.
It is easy to see that $S''$ must contain the same commands
as the output of the update detector, $U$, as both are simple
but reflect all changes between $\FS$ and $\FS^*$.
If so, then $S''\in\orderset{U}$ and based on \cref{simple_reorder_equiv}
we have
$S\equiv S'\eqext S''\equiv U$, and as $S\FS = \FS^* \neq \fsbroken$,
therefore $U\FS\neq\fsbroken$.
\end{proof}

@@

\begin{mydef}{Independent commands, sequences and sets.}
For two commands $\cxynv$ and $\czwmv$ we write $\cxynv\indep\czwmv$ if $\cxynv\cc\czwmv \equiv \czwnv\cc\cxynv \nequiv \cbrk$.
For two sequences or sets of commands $A$ and $B$ we write $A\indep B$ if
$\forall \alpha\in A \mbox{~and~} \forall \beta\in B: \alpha\indep\beta$.
\end{mydef}

\begin{mydef}{Reconciliation.}\label{def:reconciliation}
\begin{align*}
\reca = \{\alpha| & \\
& \alpha\in A\setminus B \\
& \wedge \\
& \forall \beta\in B\setminus A: \alpha\indep\beta\\
\} &
\end{align*}

In other words, the sequence $\reca$ we apply to $\FS_B$ 
is the largest subset of $A\setminus B$
that is independent of $B\setminus A$.
The order of commands in the sequence can be determined using the algorithm described in
\cref{ordering} as any subset of $A$ is also simple.
In a similar way, $\recb$ can be obtained by reversing $A$ and $B$.
\end{mydef}

@@

\begin{mylem}\label{unrelated_is_independent}
Two different commands that are not assertion commands are independent iff the nodes they change are unrelated:
\[ \cxynv\indep\czwmv \Leftrightarrow n\unrel m \]
if $\cxynv\neq\czwmv$ and none of $\cxy$ and $\cxw$ is $\cbb$ or $\cdd$.
\end{mylem}
\begin{proof}
This is easy to show based on the output of the software that investigates the behavior of command pairs,
and which we used to derive the \namecrefs{ax_separate_commute} in \cref{section_axioms}.
However, this proposition can also be derived from the \namecrefs{ax_separate_commute} already listed
in the following way.

We proceed in two steps. First, we note that
from \cref{ax_separate_commute,ax_separate_nobreaks} we know that
if $n\unrel m$, then $\cxynv$ and $\czwmv$ must necessarily be independent, and therefore
$ \cxynv\nindep\czwmv \wedge n\unrel m $ cannot be the case.

Next, we prove that
\[ \cxynv\indep\czwmv \wedge n\nunrel m \] is also impossible as it would lead to contradiction.
As $n\nunrel m$
we know that $n=m$, $n\descendant m$ or $m\descendant n$.

@@

\begin{myth}
If $A, B$ are simple, then $\worksc{B\cc \reca}{A,B}$,
where, to restate \cref{def:reconciliation},
% { % scope
% \renewcommand{\themydef}{\ref{def:reconciliation}}
% \begin{mydef}
\begin{align*}
\reca = \{\alpha| & \\
& \alpha\in A\setminus B \\
& \wedge \\
& \forall \beta\in B\setminus A: \alpha\indep\beta\\
\}. &
\end{align*}
% \end{mydef}
% } % scope
\end{myth}

@@

In these cases, either $m=\parent(n)$ or $n=\parent(m)$.
If $B$ has a command on $m$, then
from \cref{lemma:neighbor}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxynv$ (also in $B$),
and from \cref{simple_reorder_equiv} we know that $B'\equiv B$.
When describing our ordering algorithm in \cref{ordering} we saw
that the command on the parent path deremines whether a pair of commands
is a construction or destruction pair,
and that this, in turn, determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $m$ must be on
the same side of the command on $n$ in both sequences.

If $B$ has no command on $m$, then let $B'$ be $B$, 
but with an extra command added to it just before $\cxynv$
according to \cref{ax_child_assert,ax_parent_assert}, 
from which we know that $B'\equiv B$.
We also know that $B'$ is still minimal (but no longer simple).

@@

Going back to the construction and destruction pairs we see that the output type of the first command
is always the same, and it is either $D$ or $\empt$ depending on the relationship between $n$ and $m$.
Therefore $R=W$ and $\cqrmv=\czwmv$
as $|\setd|=|\setb|=1$. 
If $B$ originally had a command on $m$,
this is a contradiction as $\czwmv$ was not marked, but we see it must also be in $B'$ and therefore in $B$.
If $B$ had no command on $m$,
this is a contradiction because the command on $m$ in $B'$ is an assertion command, so $Z=Q=R=W$, 
but $A$ contained no assertion commands.

@@

We would like to prove that $\worksc{B\cc \reca}{A,B}$.
From the above we know that this is equivalent to
\[ \worksc{\acb\cc \bma\cc \reca}{A,B}. \]
As we already know that $\reca\indep\bma$
and that $\worksc{\acb\cc \bma}{A,B}$
(as $\worksc{B}{A,B}$),
based on \cref{indep_prefix_combine}
we only need to prove that
\[ \worksc{\acb\cc \reca}{A,B}. \]
Using \cref{r_invmove} we can rephrase this as
\[ \worksc{\reca}{\acbi\cc A,\acbi\cc B}. \]

@@

\subsection{Reconciliation is complete}

% Reconciliation is maximal
% -------------------------
% This is where we're using |D|=1
% TODO C.f. strict / liberal etc. reconciliation in references

The reconciliation algorithm defined above is also complete, that is,
it is not possible to apply any further commands from $A\setminus B$ to $\FS_B$.
To show this, we are going to prove the following \namecref{rec_is_complete}:

\begin{myth}\label{rec_is_complete}
If $A$ and $B$ are simple sequences,
and we select a sequence $S\subset \amb$ so that
it contains a command $\cxynv$ for which $\cxynv\nindep\bma$,
then $B\cc S$ either breaks all filesystems $A$ and $B$ do not break
(that is, $S$ breaks all possible $\FS_B$ replicas),
or it overrides incompatible changes introduced in the replica by $B$.
\end{myth}
Overriding changes is not desirable as in these cases the reconciler
does not have enough information to decide whether to promote
the change introduced by $A$ over that introduced by $B$ or vice versa.
These cases should be treated as conflicts and
the affected nodes or commands marked for review
by a human or a different system.
\begin{proof}
The proof is similar to the ones we have seen above.
Without loss of generality, we assume that $\cxynv$ is the first command in $S$
that is not independent of $\bma$.
If so, we can split $S$ into $S_0\cc\cxynv\cc S_1$ where $S_0\indep\bma$.
Let us also split $\bma$ into $T_0\cc\czwmv\cc T_1$ where $T_1\indep\cxynv$.
From \cref{can_move_intersection} we know that
$B\cc S \equiv \acb\cc\bma\cc S$.
We also know commands in $S_0$ and $\bma$ commute, and so this is equivalent to
$\acb\cc S_0\cc\bma\cc\cxynv\cc S_1$.
Expanding $\bma$ and swapping $T_1$ and $\cxynv$ we get
\[ \acb\cc S_0\cc T_0 \cc \czwmv \cc \cxynv\cc T_1\cc S_1. \]

First, we prove that this sequence breaks all filesystems
unless $n=m$.
Let us therefore suppose $n\neq m$ and, to use an inverse proof,
that the sequence works on $\FS$ where $A\FS\neq\fsbroken$.
If so, its prefix, $\acb\cc S_0\cc T_0 \cc \czwmv \cc \cxynv$ must also work.

We know $\cxynv\nindep\czwmv$, and so from \cref{unrelated_is_independent}, $n\nunrel m$.
As the sequence works, $\czwmv\cc\cxynv$ can only be a construction or destruction pair.
As $\czwmv$ and $T_0$ originate from $B\setminus A$ and $B$ is simple, we also know there are
no commands on $m$ in $A\cap B$ or $T_0$,
and that there are no commands on $m$ in $S_0$ as $S_0\indep\bma$.

On the one hand we therefore know that $\FS(m)$ must be of type $Z$,
but on the other, from \cref{ax_directchild_breaks,ax_directparent_breaks},
we also know that $\cxynv$ cannot be applied to $\FS$ without
changing $\FS(m)$ first.
As $A\FS\neq\fsbroken$, this means that $A$ must contain a command on $m$.
Let this command be $\cqrmv$.

Clearly $Q=Z$ (from \cref{worksinputmatch}), and from the construction and destruction
pairs we also see that $R=W$ must hold, where $W=\cchard$ or $W=\ccharb$.
As $|\setd|=1$ and $|\setb|=1$, this also means that $v_R=v_W$ and therefore
$\cqrmv = \czwmv$.
This is a contradiction as $\cqrmv\in A$, but $\czwmv\in\bma$.

The only possibility is therefore $n=m$.
As $A$ and $B$ are simple, we therefore know that these are the first commands on $n$,
and so (again from \cref{worksinputmatch}) $Z=X$.
We also know $\valvw\neq\valvy$ as otherwise the two commands would be equal and
would be included in $A\cap B$,
but this means that $\cxynv$, from $A$, overrides a change introduced 
by $\czwnv$, from $B$, which should be treated as a conflict.
\end{proof}

From the proof it is apparent that this result depends on $|\setd|=1$.

@@

TODO: next steps. multiple replicas
